---
title: "Housing Affordability Project"
author: "Lucia Rojkova et. al."
date: "`r Sys.Date()`"
output: html_document
---

## 0. Install Dependencies

In the first step we install libraries and dependencies we use in the code. This
step is not necessary if we use RStudio, since the libraries can be installed
there manually.

The installed libraries are

  - `tidyverse`: Working with data, processing data, cleaning data, statistics and
visualizations.
  - `eurostat`: R package for downloading datasets from Eurostat, so only the
dataset ID is needed and all data is downloaded automatically
  - `lubridata`: R package for parsing and reading date formats. In datasets dates
are usually in different formats and often we want to extract only a part of the
date such as year. This package makes working with different date formats
easier.

```{r setup, include=FALSE}
# Install necessary libraries

# Set CRAN repository for reproducibility
options(repos = c(CRAN = "https://cloud.r-project.org"))

# Load necessary libraries.
install.packages("tidyverse")
install.packages("eurostat")
install.packages("lubridate")

library(tidyverse)
library(eurostat)
library(lubridate)
```

## 1. Directory Setup

In this step we create 2 directories:

  - `data/raw/`: We will store all data downloaded from Eurostat here.
  - `data/processed/`: We will store the processed data here. 


We just need to ensure that `data/raw/` and `data/processed/` exist. If they
don't, create them.

```{r create-directories}
raw_dir <- "data/raw"
processed_dir <- "data/processed"

if (!dir.exists(raw_dir)) dir.create(raw_dir, recursive = TRUE)
if (!dir.exists(processed_dir)) dir.create(processed_dir, recursive = TRUE)
```

## 2. Helper Function to Download Eurostat Tables

Define a small helper function that checks for an existing CSV (dataset) in
`data/raw/`. 

If absent, it downloads the full raw table via `get_eurostat()` (using the
`eurostat` package) and saves it as `<dataset_code>_raw.csv`.

This code is useful because it automatically downloads all datasets from source
(so we know it is up-to-date and reproducible -> everyone with the code can work
with the same data as us), and we do not unnecessarily download it more times if
we already have it.

```{r download-function}
download_eurostat_raw <- function(dataset_code, raw_dir) {
  # Set raw_file to be the filename where we want to save the dataset
  raw_file <- file.path(raw_dir, paste0(dataset_code, "_raw.csv"))

  # If the file does not exist, download and save it
  if (!file.exists(raw_file)) {
    message("Downloading ", dataset_code, " from Eurostat...")

    # Download the dataset using the eurostat package
    df <- eurostat::get_eurostat(dataset_code, time_format = "date")

    # Save the data as a CSV to `raw_file`
    readr::write_csv(df, raw_file)

    message("Saved to ", raw_file)
  } else {
    # The dataset is already downloaded and we do not need to do anything.
    message(raw_file, " already exists; skipping download.")
  }

  raw_file
}
```

## 3. Download Specified Datasets

We will download the following five Eurostat datasets:

* `lfst_hhnhtych`: Number of households by household composition, number of children, and age of youngest child
* `sdg_10_10`: Purchasing power adjusted GDP per capita
* `prc_hicp_midx`: HICP - monthly data (index)
* `ilc_lvho07a`: Housing cost overburden rate by age, sex, and poverty status
* `tps00001`: Population on 1 January

```{r download-datasets}
datasets <- c(
  "lfst_hhnhtych", # Household composition by type
  "sdg_10_10", # GDP per capita (PPP)
  "prc_hicp_midx", # HICP monthly index
  "ilc_lvho07a", # Housing cost overburden rate
  "tps00001" # Population on 1 January
)

# Initialize an empty vector to store file paths
raw_paths <- character(length(datasets))

# Loop through each dataset and download it
for (i in 1:length(datasets)) {
  current_dataset <- datasets[i]

  # Download the raw Eurostat data and get the file path where it was saved
  raw_path <- download_eurostat_raw(
    dataset_code = current_dataset,
    raw_dir = raw_dir
  )

  # Store this file path in our results vector
  raw_paths[i] <- raw_path
}

raw_paths
```

After running the above code, you should see CSV files in `data/raw/` such as:

```
data/raw/lfst_hhnhtych_raw.csv
data/raw/sdg_10_10_raw.csv
data/raw/prc_hicp_midx_raw.csv
data/raw/ilc_lvho07a_raw.csv
data/raw/tps00001_raw.csv
```

## 4. Process HICP: Compute Annual Averages

The HICP index contains the prices of products (see the dataset documentation).
This can be used to gauge the inflation and the general increase of prices. It
is a good control variable to account for inflation. This can be used to gauge
the inflation and the general increase of prices. It is a good control variable
to account for inflation.

However, there is a problem with the dataset because it includes data for every
month. We want to work only with years. To solve this, we take the mean of the
prices over all months in each year.

For the `prc_hicp_midx` dataset (monthly HICP index), we will:

1. Read the raw dataset from CSV
2. Filter to the “all-items” `coicop` code (`CP00`)
3. Compute an annual average of the `values` column for each country (`geo`)
4. Save the resulting processed table as `prc_hicp_annual_avg.csv` under `data/processed/`

```{r process-hicp}
# Load the downloaded data CSV from disk
hicp_raw_file <- file.path(raw_dir, "prc_hicp_midx_raw.csv")

hicp_annual <- read_csv(hicp_raw_file, show_col_types = FALSE) |>
  # Keep only the `all‐items` index (This code can be found in Eurostat
  # documentation for this particular dataset)
  filter(coicop == "CP00") |>

  # Create a new column `year` with the year value extracted from the
  # `TIME_PERIOD` column
  mutate(year = lubridate::year(TIME_PERIOD)) |>

  # Group data by country (`geo`) and year (`year`)
  group_by(geo, year) |>

  # Create a new column `hicp_annual_avg` that will will be contain the mean of
  # the `values` column for each group (geo, year)
  summarise(
    # Remove any null values before summary
    hicp_annual_avg = mean(values, na.rm = TRUE),

    # When you use `group_by(geo, year)`, you're telling R to treat your data as
    # a collection of mini-datasets, one for each unique combination of country
    # and year. Think of it like sorting a deck of cards into separate piles
    # based on suit and rank - each pile represents one group.

    # When `summarise()` processes grouped data, it has to make a decision about
    # what to do with the grouping structure afterwards. By default,
    # `summarise()` follows this rule: *it removes the last grouping variable
    # but keeps all the others.*
    #
    # In this case, since we grouped by geo and year, after `summarise()` runs:
    #
    #  - The `year` grouping gets dropped (it was the last variable)
    #  - The `geo` grouping remains active
    #  - The resulting data is still grouped by country
    #
    # When we specify .groups = "drop", you're explicitly telling R: "After you
    # finish summarizing, remove ALL grouping and give me back a regular
    # tibble."
    #
    # This prevents any surprise behaviors in subsequent operations.
    .groups = "drop"
  )

# Create a new filename in the `processed_dir` for the processed data
processed_file <- file.path(processed_dir, "prc_hicp_annual_avg.csv")

# Save the processed data to disk
write_csv(hicp_annual, processed_file)

message("Processed HICP annual averages written to ", processed_file)
```

After executing the above, you will find:

```
data/processed/prc_hicp_annual_avg.csv
```

which contains columns:

* `geo` (country code)
* `year` (calendar year)
* `hicp_annual_avg` (average of monthly HICP values for that year)

```{r load-and-combine-raw, message=FALSE, warning=FALSE}
# Load all CSV data from disk
lfst_raw <- read_csv("data/raw/lfst_hhnhtych_raw.csv", show_col_types = FALSE)
gdp_raw <- read_csv("data/raw/sdg_10_10_raw.csv", show_col_types = FALSE)
hicp_annual <- read_csv("data/processed/prc_hicp_annual_avg.csv", show_col_types = FALSE)
hcob_raw <- read_csv("data/raw/ilc_lvho07a_raw.csv", show_col_types = FALSE)
pop_raw <- read_csv("data/raw/tps00001_raw.csv", show_col_types = FALSE)
```

## 2. Process Each Dataset into (geo, year, value)

In this section, we process every dataset so that it has columns

 - `geo`: Code of the country (e.g. `SK`, `CZ`, `DE`, ...)
 - `year`: The year when the data was collected
 - "value": The actual value of the particular index/data point we are
 interested in. The name of the value will be specific and descriptive for each
 individual index.

### 2.1 One-Person Households (`lfst_hhnhtych`)

In the `lfst_hhnhtych` dataset, the column for household composition is
`hhtyp`. 

We know that `"HHC_22"` corresponds to “one-person” households (this can be
found in the documentation of the dataset on the Eurostat website). 


```{r process-lfst}
lfst_proc <-
  lfst_raw |>

  # Create a new column `year` with the year value extracted from the
  # `TIME_PERIOD` column
  mutate(year = year(TIME_PERIOD)) |>

  # Keep only one-person households (hhtyp == "HHC_22")
  filter(hhcomp == "HHC_22") |>

  # Group data by country (`geo`) and year (`year`)
  group_by(geo, year) |>

  # Get the mean number of one-person households for a year in every country
  summarise(
    one_person_hh = mean(values, na.rm = TRUE),
    .groups = "drop"
  )
```

### 2.2 GDP per Capita (`sdg_10_10`)

This dataset is already annual. We extract `year` from `TIME_PERIOD` and rename
`values` to `gdp_per_capita`.

```{r process-gdp}
gdp_proc <-
  gdp_raw |>

  # Create a new column `year` with the year value extracted from the
  # `TIME_PERIOD` column
  mutate(year = year(TIME_PERIOD)) |>

  # Group data by country (`geo`) and year (`year`)
  group_by(geo, year) |>

  # Get the mean GDP per-capita for each year in every country
  summarise(
    gdp_per_capita = mean(values, na.rm = TRUE),
    .groups = "drop"
  )
```

### 2.3 HICP Annual Averages (`prc_hicp_midx`)

We already computed annual averages in `hicp_annual`, with columns `(geo, year, hicp_annual_avg)`. 
Rename for consistency.

```{r process-hicp-rename}
hicp_proc <-
  hicp_annual |>
  # Just rename the column for the index
  rename(hicp_index = hicp_annual_avg)
```

### 2.4 Housing Cost Overburden Rate (`ilc_lvho07a`)

We want the overall average (all ages, both sexes, all income situations).
Typically, the raw file contains dimensions:

* `age` (e.g., `"TOTAL"` for all ages)
* `sex` (e.g., `"T"` for total/both)
* `incgrp` (e.g., `"TOT"` for total population)

These values can be found documented on the Eurostat website.

```{r process-hcob}
hcob_proc <-
  hcob_raw |>

  # Create a new column `year` with the year value extracted from the
  # `TIME_PERIOD` column
  mutate(year = year(TIME_PERIOD)) |>

  # Keep only overall totals
  filter(
    age == "TOTAL",
    sex == "T",
    incgrp == "TOTAL"
  ) |>

  # Group data by country (`geo`) and year (`year`)
  group_by(geo, year) |>

  # Get the mean overburden rate for each year in every country
  summarise(
    hc_overburden = mean(values, na.rm = TRUE),
    .groups = "drop"
  )
```

### 2.5 Population on 1 January (`tps00001`)

Extract `year` from `TIME_PERIOD` and rename `values` to `population`.

```{r process-pop}
pop_proc <-
  pop_raw |>

  # Create a new column `year` with the year value extracted from the
  # `TIME_PERIOD` column
  mutate(year = year(TIME_PERIOD)) |>

  # Group data by country (`geo`) and year (`year`)
  group_by(geo, year) |>

  # Get the mean Population for each year in every country
  summarise(
    population = mean(values, na.rm = TRUE),
    .groups = "drop"
  )
```

## 3. Combine All Indices into One Data Frame

We now perform a join all datasets on `(geo, year)` so that every country-year
pair appears, even if one dataset is missing for that pair. After joining, each
row will have:

* `one_person_hh`
* `gdp_per_capita`
* `hicp_index`
* `hc_overburden`
* `population`

```{r combine-all}
combined_df <-
  lfst_proc |>
  full_join(gdp_proc, by = c("geo", "year")) |>
  full_join(hicp_proc, by = c("geo", "year")) |>
  full_join(hcob_proc, by = c("geo", "year")) |>
  full_join(pop_proc, by = c("geo", "year")) |>

  # Normalize one‐person households by population:
  mutate(
    one_person_hh_per_capita = one_person_hh / population
  ) |>
  arrange(geo, year)

# Inspect the first few rows
print(combined_df, n = 10)
```

### 3.1 Save the Combined Table

Save the final processed and cleaned dataset to disk, so that we can just load
it and continue from this point.

```{r save-combined}
write_csv(combined_df, "data/processed/combined_indices.csv")
message("Saved combined indices to data/processed/combined_indices.csv")
```


```{r}
# Assuming combined_df is already in your environment
combined_df |>
  # Remove rows where one_person_hh is NA
  filter(
    geo == "DE" |
      geo == "CZ" |
      geo == "PL" |
      geo == "SK" |
      geo == "EU28"
  ) |>
  filter(!is.na(one_person_hh_per_capita)) |>
  ggplot(aes(x = year, y = one_person_hh_per_capita, color = geo, group = geo)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Share of One-Person Households Over Time by Country",
    x = "Year",
    y = "Number of One-Person Households (in 1,000s)",
    color = "Country"
  ) +
  theme_minimal()
```

## Data Exploration

This section examines how the distributions of our key indicators changed before (≤ 2021) and after (≥ 2022) the outbreak of the Ukraine war (Feb 24 2022). We will generate, for each numeric variable in **`combined_df`**:

1. A QQ plot with pre‐war quantiles on the x‐axis and post‐war quantiles on the y‐axis.
2. An overlaid histogram + density plot, with pre‐war and post‐war distributions distinguished by color/ fill.

Throughout, we:

- Define **pre‐war** as all observations where `year ≤ 2021`  
- Define **post‐war** as all observations where `year ≥ 2022`  
- Drop `NA` values on a per‐variable basis  
- Use **tidyverse** and the native pipe `|>` exclusively  
- Keep each plot in its own code chunk with a brief description  

```{r load-packages-and-data, echo=TRUE}
# If combined_df is not yet in memory, read it:
# (Adjust the path if necessary; this assumes you have already run the preprocessing step.)
if (!exists("combined_df")) {
  combined_df <- read_csv("data/processed/combined_indices.csv", show_col_types = FALSE)
}

# Preview the six numeric variables we'll explore:
# - one_person_hh
# - gdp_per_capita
# - hicp_index
# - hc_overburden
# - population
# - one_person_hh_per_capita
glimpse(combined_df)
```

---

### QQ Plots

Below are QQ plots for each variable. In each plot:

* The **x‐axis** shows quantiles of the **pre‐war** sample (`year ≤ 2021`).
* The **y‐axis** shows quantiles of the **post‐war** sample (`year ≥ 2022`).
* We compute quantiles at 100 evenly‐spaced probabilities (`probs = seq(0,1, length.out = 100)`), after dropping `NA`.

#### 1. QQ Plot: One-Person Households (`one_person_hh`)

*This QQ plot compares the distribution of the total number of one‐person households before vs. after the war.*

```{r qq-one_person_hh, fig.width=6, fig.height=5, echo=TRUE}
# Extract non‐NA values pre‐war and post‐war
pre_op <- combined_df |>
  filter(year <= 2021) |>
  pull(one_person_hh_per_capita) |>
  na.omit()
post_op <- combined_df |>
  filter(year >= 2022) |>
  pull(one_person_hh_per_capita) |>
  na.omit()

# Compute matching quantiles (100 points)
probs <- seq(0, 1, length.out = 100)
q_pre <- quantile(pre_op, probs = probs, na.rm = TRUE)
q_post <- quantile(post_op, probs = probs, na.rm = TRUE)

# Build a tibble of quantiles and plot
tibble(pre = q_pre, post = q_post) |>
  ggplot(aes(x = pre, y = post)) +
  geom_point(alpha = 0.6) +
  geom_abline(linetype = "dashed", color = "gray50") +
  labs(
    title = "QQ Plot: One-Person Households (pre-war vs. post-war)",
    x = "Pre-war Quantiles (2013–2021)",
    y = "Post-war Quantiles (2022–2024)"
  ) +
  theme_minimal()
```

#### 2. QQ Plot: GDP per Capita (`gdp_per_capita`)

*This QQ plot compares the distribution of GDP per capita (PPP) before vs. after the war.*

```{r qq-gdp_per_capita, fig.width=6, fig.height=5, echo=TRUE}
pre_gdp <- combined_df |>
  filter(year <= 2021) |>
  pull(gdp_per_capita) |>
  na.omit()
post_gdp <- combined_df |>
  filter(year >= 2022) |>
  pull(gdp_per_capita) |>
  na.omit()

q_pre <- quantile(pre_gdp, probs = probs, na.rm = TRUE)
q_post <- quantile(post_gdp, probs = probs, na.rm = TRUE)

tibble(pre = q_pre, post = q_post) |>
  ggplot(aes(x = pre, y = post)) +
  geom_point(alpha = 0.6) +
  geom_abline(linetype = "dashed", color = "gray50") +
  labs(
    title = "QQ Plot: GDP per Capita (pre-war vs. post-war)",
    x = "Pre-war Quantiles (2013–2021)",
    y = "Post-war Quantiles (2022–2024)"
  ) +
  theme_minimal()
```

#### 3. QQ Plot: HICP Index (`hicp_index`)

*This QQ plot compares the distribution of the annual HICP index (all items) before vs. after the war.*

```{r qq-hicp_index, fig.width=6, fig.height=5, echo=TRUE}
pre_hicp <- combined_df |>
  mutate(hicp_index_log = log(hicp_index)) |>
  filter(year <= 2021) |>
  pull(hicp_index_log) |>
  na.omit()
post_hicp <- combined_df |>
  mutate(hicp_index_log = log(hicp_index)) |>
  filter(year >= 2022) |>
  pull(hicp_index_log) |>
  na.omit()

q_pre <- quantile(pre_hicp, probs = probs, na.rm = TRUE)
q_post <- quantile(post_hicp, probs = probs, na.rm = TRUE)

tibble(pre = q_pre, post = q_post) |>
  ggplot(aes(x = pre, y = post)) +
  geom_point(alpha = 0.6) +
  geom_abline(linetype = "dashed", color = "gray50") +
  labs(
    title = "QQ Plot: HICP Index (pre-war vs. post-war)",
    x = "Pre-war Quantiles (2013–2021)",
    y = "Post-war Quantiles (2022–2024)"
  ) +
  theme_minimal()
```

#### 4. QQ Plot: Housing Cost Overburden Rate (`hc_overburden`)

*This QQ plot compares the distribution of the housing cost overburden rate (all ages, sexes, income groups) before vs. after the war.*

```{r qq-hc_overburden, fig.width=6, fig.height=5, echo=TRUE}
pre_hc <- combined_df |>
  filter(year <= 2021) |>
  pull(hc_overburden) |>
  na.omit()
post_hc <- combined_df |>
  filter(year >= 2022) |>
  pull(hc_overburden) |>
  na.omit()

q_pre <- quantile(pre_hc, probs = probs, na.rm = TRUE)
q_post <- quantile(post_hc, probs = probs, na.rm = TRUE)

tibble(pre = q_pre, post = q_post) |>
  ggplot(aes(x = pre, y = post)) +
  geom_point(alpha = 0.6) +
  geom_abline(linetype = "dashed", color = "gray50") +
  labs(
    title = "QQ Plot: Housing Cost Overburden Rate (pre-war vs. post-war)",
    x = "Pre-war Quantiles (2013–2021)",
    y = "Post-war Quantiles (2022–2024)"
  ) +
  theme_minimal()
```

#### 5. QQ Plot: Population (`population`)

*This QQ plot compares the distribution of total population (on January 1) before vs. after the war.*

```{r qq-population, fig.width=6, fig.height=5, echo=TRUE}
pre_pop <- combined_df |>
  filter(year <= 2021) |>
  pull(population) |>
  na.omit()
post_pop <- combined_df |>
  filter(year >= 2022) |>
  pull(population) |>
  na.omit()

q_pre <- quantile(pre_pop, probs = probs, na.rm = TRUE)
q_post <- quantile(post_pop, probs = probs, na.rm = TRUE)

tibble(pre = q_pre, post = q_post) |>
  ggplot(aes(x = pre, y = post)) +
  geom_point(alpha = 0.6) +
  geom_abline(linetype = "dashed", color = "gray50") +
  labs(
    title = "QQ Plot: Population (pre-war vs. post-war)",
    x = "Pre-war Quantiles (2013–2021)",
    y = "Post-war Quantiles (2022–2024)"
  ) +
  theme_minimal()
```

#### 6. QQ Plot: One-Person Households per Capita (`one_person_hh_per_capita`)

*This QQ plot compares the distribution of one‐person households per capita before vs. after the war.*

```{r qq-one_person_hh_per_capita, fig.width=6, fig.height=5, echo=TRUE}
pre_op_pc <- combined_df |>
  filter(year <= 2021) |>
  pull(one_person_hh_per_capita) |>
  na.omit()
post_op_pc <- combined_df |>
  filter(year >= 2022) |>
  pull(one_person_hh_per_capita) |>
  na.omit()

q_pre <- quantile(pre_op_pc, probs = probs, na.rm = TRUE)
q_post <- quantile(post_op_pc, probs = probs, na.rm = TRUE)

tibble(pre = q_pre, post = q_post) |>
  ggplot(aes(x = pre, y = post)) +
  geom_point(alpha = 0.6) +
  geom_abline(linetype = "dashed", color = "gray50") +
  labs(
    title = "QQ Plot: One-Person Households per Capita (pre-war vs. post-war)",
    x = "Pre-war Quantiles (2013–2021)",
    y = "Post-war Quantiles (2022–2024)"
  ) +
  theme_minimal()
```

---

### Histogram + Density Comparisons

In the plots below, each panel shows:

* An overlaid **histogram** of the chosen variable, with `y = ..density..`.
* An overlaid **density curve** (same scale) for pre‐war vs. post‐war.
* We use `fill = period` (semi‐transparent) for the histogram and `color = period` for the density line.
* We bin each histogram into 30 bins. Adjust `bins` if you prefer more/fewer bins.

#### 1. Histogram + Density: One-Person Households (`one_person_hh`)

```{r hist-one_person_hh, fig.width=6, fig.height=5, echo=TRUE}
# Prepare a data frame with a new 'period' column
hist_op_df <- combined_df |>
  filter(!is.na(one_person_hh)) |>
  mutate(period = if_else(year <= 2021, "Pre-war", "Post-war"))

ggplot(hist_op_df, aes(x = one_person_hh, fill = period)) +
  geom_histogram(aes(y = after_stat(density)),
    alpha = 1, position = "identity", bins = 30
  ) +
  geom_density(aes(color = period), size = 1, alpha = 0.4, outline.type = ) +
  labs(
    title = "Distribution: One-Person Households (pre-war vs. post-war)",
    x = "One-Person Households",
    y = "Density",
    fill = "Period",
    color = "Period"
  ) +
  theme_minimal()
```

#### 2. Histogram + Density: GDP per Capita (`gdp_per_capita`)

```{r hist-gdp_per_capita, fig.width=6, fig.height=5, echo=TRUE}
hist_gdp_df <- combined_df |>
  filter(!is.na(gdp_per_capita)) |>
  mutate(period = if_else(year <= 2021, "Pre-war", "Post-war"))

ggplot(hist_gdp_df, aes(x = gdp_per_capita, fill = period)) +
  geom_histogram(aes(y = ..density..),
    alpha = 0.4, position = "identity", bins = 30
  ) +
  geom_density(aes(color = period), size = 1) +
  labs(
    title = "Distribution: GDP per Capita (pre-war vs. post-war)",
    x = "GDP per Capita (PPP)",
    y = "Density",
    fill = "Period",
    color = "Period"
  ) +
  theme_minimal()
```

#### 3. Histogram + Density: HICP Index (`hicp_index`)

```{r hist-hicp_index, fig.width=6, fig.height=5, echo=TRUE}
hist_hicp_df <- combined_df |>
  filter(!is.na(hicp_index)) |>
  mutate(period = if_else(year <= 2021, "Pre-war", "Post-war"))

ggplot(hist_hicp_df, aes(x = hicp_index, fill = period)) +
  geom_histogram(aes(y = ..density..),
    alpha = 0.4, position = "identity", bins = 30
  ) +
  geom_density(aes(color = period), size = 1) +
  labs(
    title = "Distribution: HICP Index (pre-war vs. post-war)",
    x = "Annual HICP Index (All Items)",
    y = "Density",
    fill = "Period",
    color = "Period"
  ) +
  theme_minimal()
```

#### 4. Histogram + Density: Housing Cost Overburden Rate (`hc_overburden`)

```{r hist-hc_overburden, fig.width=6, fig.height=5, echo=TRUE}
hist_hc_df <- combined_df |>
  filter(!is.na(hc_overburden)) |>
  mutate(period = if_else(year <= 2021, "Pre-war", "Post-war"))

ggplot(hist_hc_df, aes(x = hc_overburden, fill = period)) +
  geom_histogram(aes(y = ..density..),
    alpha = 0.4, position = "identity", bins = 30
  ) +
  geom_density(aes(color = period), size = 1) +
  labs(
    title = "Distribution: Housing Cost Overburden Rate (pre-war vs. post-war)",
    x = "Housing Cost Overburden Rate (%)",
    y = "Density",
    fill = "Period",
    color = "Period"
  ) +
  theme_minimal()
```

#### 5. Histogram + Density: Population (`population`)

```{r hist-population, fig.width=6, fig.height=5, echo=TRUE}
hist_pop_df <- combined_df |>
  filter(!is.na(population)) |>
  mutate(period = if_else(year <= 2021, "Pre-war", "Post-war"))

ggplot(hist_pop_df, aes(x = population, fill = period)) +
  geom_histogram(aes(y = ..density..),
    alpha = 0.4, position = "identity", bins = 30
  ) +
  geom_density(aes(color = period), size = 1) +
  labs(
    title = "Distribution: Population (pre-war vs. post-war)",
    x = "Population (Jan 1)",
    y = "Density",
    fill = "Period",
    color = "Period"
  ) +
  theme_minimal()
```

#### 6. Histogram + Density: One-Person Households per Capita (`one_person_hh_per_capita`)

```{r hist-one_person_hh_per_capita, fig.width=6, fig.height=5, echo=TRUE}
hist_op_pc_df <- combined_df |>
  filter(!is.na(one_person_hh_per_capita)) |>
  mutate(period = if_else(year <= 2021, "Pre-war", "Post-war"))

ggplot(hist_op_pc_df, aes(x = one_person_hh_per_capita, fill = period)) +
  geom_histogram(aes(y = ..density..),
    alpha = 0.4, position = "identity", bins = 30
  ) +
  geom_density(aes(color = period), size = 1) +
  labs(
    title = "Distribution: One-Person Households per Capita (pre-war vs. post-war)",
    x = "One-Person Households per Capita",
    y = "Density",
    fill = "Period",
    color = "Period"
  ) +
  theme_minimal()
```

---

**Notes on Implementation**:

1. **Defining “pre‐war” and “post‐war”**
   We split by `year <= 2021` vs. `year >= 2022`. These year‐based splits approximate “before” and “after” Feb 24 2022, since our data are annual.

2. **Dropping NA’s**
   Each chunk calls `filter(!is.na(var))` before plotting. This ensures that missing values do not interfere with quantile calculations or density estimates.

3. **Quantile Calculation for QQ plots**
   We take 100 evenly‐spaced quantile probabilities (`seq(0,1,length.out=100)`). For each variable, we compute `quantile(pre_values, probs)`, `quantile(post_values, probs)`, then make a simple scatter‐plot of `(pre_quantile, post_quantile)` with a 45° reference line.

4. **Histogram + Density**

   * We request `geom_histogram(aes(y=..density..))` with `alpha=0.4` so the density curves (plotted using `geom_density(aes(color=period))`) align correctly on the density scale.
   * We use `bins = 30` uniformly, but you can adjust that if you want finer or coarser bin‐widths.

5. **Code Organization**

   * Each plot has its own named chunk (e.g., `qq-one_person_hh`, `hist-gdp_per_capita`), with a brief descriptive caption above.
   * We rely on base `tidyverse` and the native pipe (`|>`), so this is compatible with R 4.3.0+.

With these plots in place, you can visually assess whether—and how—the distributions of each indicator shifted after the onset of the Ukraine war.
